---
title: "STAT 628 Credit Risk Project: Installment 1"
subtitle: "Spring 2026"
author: "Shengyao Ye"
date: last-modified
format: pdf
geometry:
  - margin=1in
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| label: setup
#| echo: false
knitr::opts_chunk$set(
    echo = TRUE, tidy = TRUE,
    tidy.opts = list(width.cutoff=80, arrow=TRUE, indent=2),
    fig.height = 3.375, fig.width = 6
)
```

## Problem Introduction

You have been asked to help a local credit union assess the credit risk
of potential borrowers who have applied for small business loans. The
credit union tends to attract borrowers who have started small local
business like cafes, dry cleaners, or bodegas but lack extensive credit
histories or background. Typically, these borrowers, having exhausted
most of their personal funds to start their businesses, seek short-term
loans to help keep their businesses running. However, their lack of
credit background makes these entrepreneurs too risky for large national
banks. And so these potential borrowers have turned to the local credit
union and applied for a loan.

The credit union issues loans with 12 month terms. That is, it expects
to recover the principal and interest within one year. The principals
range from about \$10,000 to \$2 million. To obtain payments from the
borrower, the credit union garnishes a fixed percentage of the credit
card transactions processed by the borrower. The credit union extracts
this percentage until it recovers not only the principal but also the
interest. That is, each month the lender take a cut from all credit card
transactions made by the business. The percentage is fixed at the time
that the loan originates and is set in anticipation that the borrower
can pay off the loan in 12 months.

Of course, there is monthly variation in credit card transactions: in
some months, businesses might have a large number of sales and in some
months, they might not have many sales. Such fluctuations contribute to
the risk the credit union assumes when it lends money. If a business
enjoys strong sales for several months, the credit union may recover its
money (plus interest) quickly. On the other hand, if a business
experiences a prolonged period of reduced sales, the credit union is
unlikely to recover its money within the year.

### An example

As an example, suppose that the owner of a new cafe obtains a \$20,000
loan from the credit union. With interest of 20%, the cafe owner must
repay a total of \$24,000 to the credit union within 12 months. To
ensure on-time repayment, the credit union wants to recover \$2,000 per
month by garnishing the cafe's credit card transactions. When the cafe
owner applied for the loan, they reported that, on average, they
processed \$15,000 per month in credit card transactions. Based on this
information, the credit union decided to garnish
$2000/15000 \times 100\% \approx 13.3\%$ of all credit card transactions
processed by the cafe.

If the cafe has a strong month of sales and processes \$22,500 in credit
card transactions, the credit union will receive \$3,000, putting the
cafe ahead of its payment schedule. On the other hand, if the cafe
experiences a slow month and only processes \$12,000 in transactions,
the credit union will receiving only \$1,600, putting the cafe behind
its payment schedule.

## Your task

To predict the riskiness of lending money to new potential borrowers,
the credit union collected data about loans they have given in the past.
You have received a dataset derived from a random sample of these past
loans containing three variables:

-   The total amount owed to the credit union (`TotalAmtOwed`). This
    amount reflects the principal of the loan as well as all interest.
    The credit union wants to recover this amount from the borrower at
    the end of the 12 months.

-   The total amount that was actually paid back to the credit union
    after six months (`AmtRepaid`). This is the actual amount of money
    that credit union recovered after six months by garnishing the
    borrower's credit card transactions.

-   A new metric that the credit union created called **P**erformance
    **R**atio at **S**ix **M**onths (`PRSM`). The PRSM score is computed
    as $$
    \text{PRSM} = 2 \times \frac{\text{Amount repaid at 6 months}}{\text{Total amount owed}}.
    $$

## Grading

Please refer to the grading rubric on Canvas. To earn all full points on
a particular question, you must (i) write **concise**, narrative answers
to each prompt using complete sentences free from grammatical or
spelling errors and (ii) include working code to produce the requested
visualizations or numerical summaries. The instructor and TA will run
your submitted code and compare it to the visualizations and answers
submitted. If they are unable to reproduce your submitted answers for a
particular question, the maximum number of points you can receive is
half of the points for that question.

\newpage

## Problem 1

### 1(a)

Load your data set into `R` (say) as well as any and all necessary
libraries (e.g., `tidyverse`).

*Solution*: <!-- Start writing solution here -->

```{r}
#| label: Q1a
# Your code goes here!
library(tidyverse)
df <- read.csv("/Users/menghanyu/Desktop/STAT 628/Module 1/installment1_group1.csv") 
head(df)
```

### 1(b)

Using appropriate numerical and graphical tools, summarize the
distributions of each of the 3 variables. Comment, in particular, on
whether each variable appears to be normally distributed or not.

*Solution*: \<!--

**TotalOwed:** The histogram shows a roughly uniform or multi-modal
distribution rather than a bell shape. The Q-Q plot deviates
significantly from the straight line at both tails (S-shape). Therefore,
TotalOwed does not appear to be normally distributed.

**AmtRepaid:** The histogram is clearly right-skewed, with a long
tail extending to the right. The mean (\$86,714) is greater than the
median (\$80,234), and the Q-Q plot shows a distinct curvature. It is
not normally distributed.

**PRSM:** The histogram is symmetric and bell-shaped. The mean (0.8210)
and median (0.8249) are very close. Additionally, the points on the Q-Q
plot align closely with the diagonal line. Thus, PRSM appears to be
normally distributed --\>

```{r}
#| label: Q1b-totalowed
# Your code for goes here.
# You may need to use multiple code blocks. 
#(Numerical Summary)
summary(df)

#(Graphical Summary)
par(mfrow = c(2, 3))

# 1: TotalOwed
hist(df$TotalOwed, main = "Hist: TotalOwed", xlab = "Total Owed", col = "skyblue")
qqnorm(df$TotalOwed, main = "QQ: TotalOwed"); qqline(df$TotalOwed, col = "red")

# 2: AmtRepaid 
hist(df$AmtRepaid, main = "Hist: AmtRepaid", xlab = "Amt Repaid", col = "lightgreen")
qqnorm(df$AmtRepaid, main = "QQ: AmtRepaid"); qqline(df$AmtRepaid, col = "red")

# 3: PRSM
hist(df$PRSM, main = "Hist: PRSM", xlab = "PRSM Score", col = "salmon")
qqnorm(df$PRSM, main = "QQ: PRSM"); qqline(df$PRSM, col = "red")

par(mfrow = c(1, 1))
```

### 1(c)

Remember that the credit union sets the garnishment percentage with the
expectation that it can recover the total amount owed in 12 months.
Accordingly, the credit union expects to have received about half of the
total amount owed after six months. The credit union regards any loan
for which the borrower has repaid less than half of the total amount
owed in six months as risky.\
Based on your data, what is the probability that a randomly selected
loan will be risky?

*Solution*: <!-- Start writing solution here 
Based on the sample data, the probability that a randomly selected loan will be risky is 86.8% -->

```{r}
#| label: Q1c
# Put any code you used for question 1(c) here. 
# If you didn't need to use any code, you can delete the codeblock.

is_risky <- df$AmtRepaid < (0.5 * df$TotalOwed)

risk_prob <- mean(is_risky)

print(paste("Probability of risky loan:", risk_prob))
```

## Problem 2

The credit union is concerned that loans with higher principal (and
therefore, higher total amounts owed) are more likely to be riskier than
loans with lower principal (and therefore, lower total amounts owed).
That is, the credit union is concerned that borrowers who owe more money
to the credit union are less likely to pay off their debt on time than
borrowers who owe less money. In this problem, you will consider the
relationship between two variables measured in your data set, the total
amount owed to the credit union (`TotalAmtOwed`) and the amount repaid
at six months (`AmtRepaid`).

### 2(a)

Visualize the relationship between `TotalAmtOwed` and `AmtRepaid`. Do
the variables appear to be independent or dependent? If they appear to
be dependent, comment on the relationship between them.

*Solution*: <!-- your solution goes here -->

```{r}
#| label: Q2a
# Your code for question 2(a) goes here
```

### 2(b)

If the amount garnished from the borrower's credit card transactions
accumulates at a rate which will ensure full repayment in 12 months,
approximately what would you expect the slope and intercept of the least
squares line to be when regressing `AmtRepaid` ($Y$) on `TotalAmtOwed`
($X$)?

*Solution*: <!-- your solution goes here -->

### 2(c)

What is the slope and intercept of the least squares line fitted to your
data? If these estimates differ from your expectations in part (b),
comment briefly on the source of these differences. Assuming that the
data adhere to the simple linear model assumptions, briefly comment on
the degree to which your estimates differ from your expectations and
explain why such differences might occur.

*Solution*: <!-- your solution goes here -->

```{r}
#| label: Q2c
# Your code for question 2(c) goes here
```

### 2(d)

Suppose that the credit union has lent money to two borrowers: Aldo who
owes a total of \$90,000 and Becky who owes \$20,000. Assuming that the
data adhere to the linear model assumptions, compute 95% prediction
intervals for the amounts that the credit union will recover from Aldo
and Becky in six months.

*Solution*: <!-- your solution here-->

```{r }
#| label: Q2d-chunk1
# Your code for question 2(d) goes here
# Feels free to use as many code chunks as you need
```

### 2(e)

Do the data used to estimate the least squares regression line in 2(c)
adhere to the linear model assumptions? If they do, briefly interpret
the value of the slope $\beta_{1}$ and the $R^{2}$. If they do not,
describe which assumptions are violated and how. You should use
appropriate diagnostic tools to justify your answer.

*Solution*: <!-- your solution goes here -->

```{r}
#| label: Q2e-chunk1
# Your code for question 2(d) goes here
# Feels free to use as many code chunks as you need
```

## Problem 3

The credit union believes that its new metric, PRSM, can be used to
quantify the riskiness of loans. Recall from Problem 2 that the credit
union is concerned that the more a borrower owes, the less likely they
are to pay the loan off on time. In this problem, you will consider the
relationship between two variables measured in your data set, the total
amount owed to the credit union (`TotalAmtOwed`) and the PRSM score
(`PRSM`).

### 3(a)

Visualize the relationship between `TotalAmtOwed` and `PRSM`. Do the
variables appear to be independent or dependent? If they appear to be
dependent, comment on the relationship between them.

*Solution*: <!-- your solution goes here -->
By visualizing the relationship between `TotalAmtOwed` and `PRSM`. the variables appear to be dependent. From the regression line, it can be found that as `TotalAmtOwed` increases,`PRSM` also slightly increases accordingly. However, the overall increase was not significant. Moreover, the scattered points are distributed relatively scattered around the regression line. From this, it can be inferred that `TotalAmtOwed` cannot well reflect the repayment performance, and the repayment ability is also affected by other factors.
```{r}
#| label: Q3a
# Your code for question 3(a) goes here
ggplot(df, aes(x = TotalOwed, y = PRSM)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "PRSM vs Total Amount Owed",
       x = "Total Amount Owed",
       y = "PRSM")
```

### 3(b)

If the amount garnished from the borrower's credit card transactions
accumulates at a rate which will ensure full repayment in 12 months,
approximately what would you expect the slope and intercept of the least
squares line when regressing `PRSM` ($Y$) on `TotalAmtOwed` ($X$) to be?

*Solution*: <!-- your solution goes here -->
According to the question, in the most expect situation, half of the `TotalAmtOwed` should be fully repaid within six months. That is to say, in the expect situation, regardless of the result of the `TotalAmtOwed`, the `PRSM` should remain at 1. So the expected intercept is 1 and the expected slope is 0.
$$
PRSM = 2 \times \frac{\text{AmtRepaid}}{\text{TotalAmtOwed}} = 2 \times \frac{0.5 \times \text{TotalAmtOwed}}{\text{TotalAmtOwed}}=1
$$
### 3(c)

What is the slope and intercept of the least squares line fitted to your
data? Assuming that the data adhere to the linear model assumptions,
briefly comment on the degree to which your estimates differ from your
expectations and explain why such differences might occur.

*Solution*: <!-- your solution goes here -->
According to the result of the code operation, it can be seen that the slope obtained after fitting is $0.7189$, and the intercept is $5.047*10^{-7}$. Based on the second question, we calculated that the expected slope is $0$and the intercept is $1$. So we can see that the intercept in the fitting situation is less than $1$, indicating that the borrower's repayment progress within six months is far lower than expected. Although the p-value of the slope in the code calculation indicates a statistically significant relationship between 'TotalAmtOwed' and 'PRSM', due to the extremely small slope, it suggests that 'TotalAmtOwed' has a very limited impact on the repayment progress. There are numerous factors in reality that can cause the outcome to deviate from the expected value, such as individual business conditions and external environmental factors.
```{r}
#| label: Q3c
# Your code for question 3(c) goes here
prsm_totalAmtOwed_model <- lm(PRSM ~ TotalOwed, data = df)
summary(prsm_totalAmtOwed_model)
```

### 3(d)

Suppose that the credit union has lent money to two borrowers: Aldo who
owes a total of \$90,000 and Becky who owes \$20,000. Assuming that data
adhere to the linear model assumptions, compute 95% prediction intervals
for the PRSM scores of Aldo's and Becky's loans. Comment on the
differences in the standard errors of both predictions.

*Solution*: <!-- start typing your solution here -->
The 95% prediction intervals for the PRSM scores of Aldo's is $(0.464,1.065)$ for Becky's is $(0.428,1.030)$And the widths of the predicted intervals for the two of them are 0.601 and 0.602 respectively. Moreover, based on the previous question, it can be known that the average value of TotalAmtOwed in this data is $202,284 $. Therefore, the loans of Aldo and Becky are both less than the average value of TotalAmtOwed, and both are greater than the minimum value of $100,471 $. Thus, the difference in their predictions is very small.
```{r}
#| label: Q3d
# Your code for question 3(d) goes here.
new_data <- data.frame(TotalOwed = c(90000, 20000))
predict(prsm_totalAmtOwed_model, new_data, interval = "prediction", level = 0.95)
summary(prsm_totalAmtOwed_model)
```

### 3(e)

Do the data used to estimate the least squares regression line in 3(c)
adhere to the linear model assumptions? If they do, briefly interpret
the value of the slope $\beta_{1}$ and the $R^{2}$. If they do not,
describe which assumptions are violated and how. You should use
appropriate diagnostic tools to justify your answer.

*Solution*: <!-- your solution goes here -->
By analyzing the relationship between the residuals and the fitted values, we obtained a correlation coefficient of $2.54*10^{-16}$, which is infinitely close to 0. It can be seen that the residuals hardly change with the predicted values. So the data used in 3(c) to estimate the least squares regression line adheres to the assumption of the linear model.
According to 3(c), we can know that the slope $\beta_{1}=5.047*10^{-7}$, that is, when TotalAmtOwed increases, PRSM has a slight increase. Since the $R^{2}=0.137$. It is indicated that only approximately 13.7% of the mutations in PRSM can be explained by TotalAmtOwed, and PRSM is also influenced by other factors
```{r}
#| label: Q3e
# Your code for question 3(e) goes here
# Feels free to use as many code chunks as you need
library(broom)
res <- residuals(prsm_totalAmtOwed_model)
fit <- fitted(prsm_totalAmtOwed_model)
cor(res, fit)
glance(prsm_totalAmtOwed_model)
```

## Problem 4

Suppose that the credit union has the opportunity to lend out enough
money to ensure its borrowers will owe a total of \$180,000. However,
the bank has two options (A and B) for lending out these funds. Under
option A, they make 2 loans, each of which will need to each pay back
\$90,000. Under option B, they make 9 loan, each of which will need to
pay back \$20,000. Due to its own credit obligations, the credit union
must recover at least \$70,000 within six months of issuing the loan.
Which option would you recommend to the credit union: lend to 2
borrowers (Option A) or lend to 9 (Option B)? Justify your choice by
estimating the probability that the credit union will meet its credit
obligations (i.e. recover at least \$70,000 in six months). You should
clearly state any assumptions you made in your analysis.

*Hint*: you might find your answers to parts 2(d) and 3(d) helpful. In
particular think about the mean and standard error and distribution of a
single prediction at different values of the predictor. Then think about
the mean, standard error, and distribution of the average of 2 and 9
predictions at different values of the predictor.

*Solution*: <!-- your solution here-->

```{r}
#| label: Q4
# Your code for Problem 4 goes here
# Use as many separate code chunks as you need
```
